Viscoelastic Kelvin-Helmholtz instability

### 2016-02-15 paper/paper.tex
improved problem formulation and fleshed out the results section of the paper

%	paper.tex	Toby Searle
%	Last modified: Fri 12 Feb 17:42:28 2016
%	using jfm template

@@

We propose that in fact, besides the shear banding instability and the
coextrusion instabilities, there is another mechanism by which a purely elastic
fluid may become unstable. Using a linear stability analysis we examine a new
Kelvin-Helmholtz like instability in both Oldroyd-B and FENE fluids. A purely
elastic fluid is found to be unstable at sufficiently high Weissenberg number.
The growth rate of the largest growing eigenmode of the instability increases
with increasing Weissenberg number at low Reynolds numbers and this behaviour
is consistent across both Oldroyd-B and FENE constituitive models.

The flow we consider is a hyperbolic tangent shaped laminar profile for the
streamwise velocity across a channel, where $y = \pm \Delta$ is the region of
shear. $y=0$ is the centreline of the channel, so that if $\Delta << 1$ the
flow can be said to be a free shear instability.  This gives a base flow
profile, 
\begin{align} 
    U(y) &= \tanh \left( y/\Delta
    \right) \coth \left( 1/\Delta \right) \nonumber\\ V &= 0 \nonumber \\
    T_{xx}(y) &= 2 Wi \left( \dy{U} \right)^{2} \nonumber \\ T_{xy}(y) &=
    \dy{U} \nonumber \\ T_{yy}(y) &= 0 \label{eq:KH_laminar_profile}
\end{align}
$U$ and $V$ are the base streamwise (x) and wall normal (y) velocities
respectively. $T$ is the base stress tensor. We use Gauss-Labatto points in the
wall-normal ($y$) direction for the base profile, and decomposed the
disturbances to this flow into Fourier modes.  
\begin{equation} 
    g(y) = \sum\limits_{n=-N}^{N} \widetilde{g}(y) e^{ikx + \lambda t} 
\end{equation}
for all disturbance variables $g = u, v, p, \tau_{i,j}$. Where $N$ is the
number of Fourier modes, $k$ is the streamwise wavenumber of the disturbance
and $\lambda$ is the growth rate of the mode. This program uses two domains of
pseudo-spectral points to give increased resolution in the centre of the
channel. The analysis was repeated using a code for which the system of points
was stretched using: 
\begin{align} 
    z_{u} &= \frac{y_{u}}{1-y_{u}} \\
    z_{b} &= \frac{y_{b}}{1+y_{b}} \\ U     &= \tanh{z/\Delta}
    \label{eq:KH_inf_profile} 
\end{align} 
This system ought to be completely insensitive to the boundary conditions at
the walls so we refer to it as 'free shear' throughout this manuscript.

The dimensionless quantities used throughout are non-dimensionalised relative
to the instability size, $\Delta$, rather than the half channel width. This is
because $\Delta$ is the length scale for the free shear instability.

We used an Oldroyd-B fluid with very low Reynolds number $\Rey \leq 1.0$ and a
small ratio between the solvent and total viscosities $\beta =
\frac{\mu_{s}}{\mu_{s}+\mu_{p}}$. This approximates a purely elastic fluid,
where the polymer contribution to the stress on the fluid is much larger  than
the solvent contribution. The Oldroyd-B Navier stokes and constitutive
equations are then: 
\begin{align} 
    Re \left[ \dt{\mathbf{v}} +
    \mathbf{v} \cdot \nabla  \mathbf{v} \right] &= - \nabla p + \beta
    \nabla^{2} \mathbf{v} + \frac{1-\beta}{Wi} \nabla \cdot \mathbf{\tau} \\
    \dot{\tau}/Wi + \overset{\nabla}\tau &= \left(\nabla \mathbf{v}\right)^{T}
    + \nabla{\mathbf{v}} 
\end{align} 
With $\mathbf{v}$ as the total (base flow and disturbance) velocity and $\tau$
as the total stress in the polymeric fluid. The FENE-CR viscoelastic fluid has
the advantage of being similarly easy to analyse using our spectral method, but
also including a finite extensibility for the polymers. This leads to a stress
which depends on a conformation tensor ($\mathbf{C}$) for the polymer dumbbells
via a non-linear spring force:

\begin{equation} 
    \mathbf{\tau} = \frac{1-\frac{3}{L^{2}}}{1 +
	\frac{L^{2}}{tr(\mathbf{C^{2}})}} \label{eq:KH_FENE_stress}
\end {equation}

\begin{figure}[h]
    \centering 
    \includegraphics[width=0.3\textwidth]{{KH_diagram}}
    \caption{Diagram of a shear layer} 
    \label{fig:KH_diagram} 
\end{figure}

Again, we used Gauss-Labatto points in the wall normal direction and Fourier
modes in the streamwise direction.

@@

At low Reynolds number, low $\beta$ and sufficiently large Weissenberg number
we observe a free shear instability across a range of streamwise disturbance
wavenumbers. This leads us to suppose that this instability might be a
plausible cause for a self-sustaining process in viscoelastic plane Couette
flow, similar to that observed by Waleffe for Newtonian flows.

We see an instability grow from around $k \sim 0.06$ and move to lower $k$ as
the Weissenberg number increases (figure \ref{fig:KH_inf_dispersions_vary_Wi}). The
height of the instability also increases with increasing Weissenberg number
(figure \ref{fig:KH_inf_disp_height_vary_Wi}).

@@

The dispersion relation saturates at $Wi \sim 50$ and remains approximately
constant with increasing Weissenberg number. The dispersion relation for the
instability remains broad in $k$ until $Wi \sim 100$ where a new eigenvector
becomes dominant.

@@

As the Reynold's number is reduced, new eigenvalues become dominant. Although
the dominant eigenvalue changes the flow is still unstable, even down to $\Rey
= 0.01$ (see figures \ref{fig:KH_inf_disp_height_vary_Re} and \ref{fig:KH_dispersions_vary_Re}).

@@

Now we use equations \ref{eq:KH_laminar_profile} and examine the unstretched
pseudospectral domain. To validate the code, we examine the high Reynolds
number regime. We see the height of the instability decrease with increasing
Weissenburg number, and this trend is stronger for smaller $\beta$ (figure
\ref{fig:KH_drag_reduction}). This is the same behaviour reported in
\cite{Azaiez1994}.

@@

\begin{figure}[h]
%   \showthe\columnwidth %use this to determine size of figures.
    \centering \includegraphics[width=0.7\textwidth]{{KH_high_Re_vary_Wi}}
    \caption{ 
	Peak growth rate against Weissenberg number at $\Rey = 1000$,
	$\Delta = 0.1$ The height of the dispersion relation is reduced for the
	Kelvin-Helmholtz instability as the Weissenberg number is increased.
	The grey dashed line is the height of the dispersion relation for the
	Newtonian fluid. The decrease in $\lambda_{r}$ corresponds to increased
	stability of the base flow, this matches the observations made in \cite{Azaiez1994}} 
    \label{fig:KH_drag_reduction} 
\end{figure}

The instability remains present under variation of $\Delta$ at low $\Delta$. At
$\Delta =0.1$ the instability is still present for $\Rey < 0.01$ far lower than
the usual threshold of $\Rey \sim 1000$ for a Newtonian Kelvin-Helmholtz
instability (figure \ref{fig:KH_walls_vary_Re}).

At high Reynolds number the instability is insensitive to the distance to the
walls when $\Delta =0.1$ (figure \ref{fig:KH_high_Re_walls_dependence}). At low
Reynolds number, the width of the channel is more important but the instability
is still present for very low $\Delta$ (figure
\ref{fig:KH_low_Re_walls_dependence}).

@@

As $\Delta$ decreases for a particular purely elastic instability, the size of
the velocities expands slightly but in the limit of very small $\Delta$ the
eigenvector appears independent of the walls (figures
\ref{fig:KH_psi_low_delta} and \ref{fig:KH_psi_high_delta}). The large width
vortices in our purely elastic case, seem to correspond to the larger
elasticity ($G~\Wi/\Rey$) eigenvectors in \citet{Azaiez1994}. Their $G=5$ case
would only correspond to $\Wi=0.5$ for $\Re=0.1$ in our purely elastic
instability, so it is a completely different, inertial instability that they
observe (\citet{Azaiez1994} figure 8). Nonetheless at least one of their
observations does apply. The fact that the velocity disturbance is essentially
zero in the center of the channel, implies that the shear layer will be less
likely to roll up. In the Newtonian instability, net velocities within the
shear layer will distort its shape, however the purely elastic instability
appears to have less of an effect on the shear layer.

@@

As the Reynold's number is reduced the walls become more important to the
instability, this makes a true $\Rey = 0$ instability difficult for us to
observe. In order to remove the dependence on the walls and recover a free
shear instability it is necessary to move them further away. This problem gets
worse at lower $\Rey$.  By introducing just a very small contribution from
inertia, we can remove the dependence on the walls. This is why all of our
results are taken for $\Rey \lesssim 0.1$ on the length scale of the
instability.

>>>>

%	paper.tex	Toby Searle
%	Last modified: Mon 15 Feb 2016 18:18:32 GMT
%	using jfm template

@@

We propose that in fact, besides the shear banding and coextrusion
instabilities, there is another mechanism by which a purely elastic fluid may
become unstable. Using a linear stability analysis we examine a new
Kelvin-Helmholtz like instability in both Oldroyd-B and FENE fluids. A purely
elastic fluid is found to be unstable at sufficiently high Weissenberg number
and this behaviour is consistent across both Oldroyd-B and FENE constituitive
models.

There are a number of ways we might formulate a shear layer problem. We are
interested in a generic instability for flows with a sharp change in the
streamwise velocity, and so large normal stresses in the fluid. Precisely
because we are searching for a generic mechanism, we will not attempt to
justify our particular choice of a hyperbolic Cosine shear layer.  As well
examining the dependence on flow properties ($\beta$, $\Wi$ and $\Rey$) of the
instability, we are also interested in the influence of boundaries so that we
might assess the relevence of the instability to common experimental and
real-life examples of flows of viscoelastic fluids - including both the plane
Couette and Oscillatory flow problems of \citet{Searle2016b} and
\citeyear{Searle2016a}. As such we consider a nearly free shear layer (with a
finite, but large, region of constant velocity between the shear layer and the
walls) as well as a free shear layer.

\begin{figure}[h]
    \centering 
    \includegraphics[width=0.3\textwidth]{{KH_diagram}}
    \caption{Diagram of a shear layer in a channel.} 
    \label{fig:KH_diagram} 
\end{figure}

The flow we consider is a hyperbolic tangent shaped laminar profile for the
streamwise velocity across a channel, where $y = \pm \Delta$ is the region of
shear. $y=0$ is the centreline of the channel, so that if $\Delta << 1$ the
flow tends to a free shear instability. This gives a base flow profile, 
\begin{align} 
    U(y) &= \tanh \left( y/\Delta
    \right) \coth \left( 1/\Delta \right) \nonumber\\ V &= 0 \nonumber \\
    T_{xx}(y) &= 2 Wi \left( \dy{U} \right)^{2} \nonumber \\ T_{xy}(y) &=
    \dy{U} \nonumber \\ T_{yy}(y) &= 0 \label{eq:KH_laminar_profile}
\end{align}
$U$ and $V$ are the base streamwise (x) and wall normal (y) velocities
respectively. $T$ is the base stress tensor. We use Gauss-Labatto points in the
wall-normal ($y$) direction for the base profile, and decomposed the
disturbances to this flow into Fourier modes.  
\begin{equation} 
    g(y) = \sum\limits_{n=-N}^{N} \widetilde{g}(y) e^{ikx + \lambda t} 
\end{equation}
for all disturbance variables $g = u, v, p, \tau_{i,j}$. Where $N$ is the
number of Fourier modes, $k$ is the streamwise wavenumber of the disturbance
and $\lambda$ is the growth rate of the mode. This program uses two domains of
pseudo-spectral points to give increased resolution in the centre of the
channel. The analysis was repeated in the limit of no boundaries on the flow by
performing a passive transform of the coordinate systems such that, 
\begin{align} 
    z &= \frac{y}{1-y},\hphantom{aaa} z>0 \\
    z &= \frac{y}{1+y},\hphantom{aaa} z<0 \\ 
    U &= \tanh{z}
    \label{eq:KH_inf_profile} 
\end{align} 
This system ought to be completely insensitive to the boundary
conditions at the walls so we refer to it as `free shear' throughout this
manuscript.

The dimensionless quantities used throughout are non-dimensionalised relative
to the instability size, $\Delta$, rather than the half channel width. This
allows us to effectively examine the dependence of the instability on $\Delta$
as we approach the free shear limit of $\Delta = 0$.

We used an Oldroyd-B fluid at low Reynolds number (typically $\Rey < 10.0$)
and a small ratio between the solvent and total viscosities $\beta =
\frac{\mu_{s}}{\mu_{s}+\mu_{p}}$. This approximates a purely elastic flow,
where the polymer contribution to the stress on the fluid is much larger  than
the solvent contribution. The Oldroyd-B Navier stokes and constitutive
equations are then: 
\begin{align} 
    Re \left[ \dt{\mathbf{v}} +
    \mathbf{v} \cdot \nabla  \mathbf{v} \right] &= - \nabla p + \beta
    \nabla^{2} \mathbf{v} + \frac{1-\beta}{Wi} \nabla \cdot \mathbf{\tau} \\
    \dot{\tau}/Wi + \overset{\nabla}\tau &= \left(\nabla \mathbf{v}\right)^{T}
    + \nabla{\mathbf{v}} 
\end{align} 
With $\mathbf{v}$ as the total (base flow and disturbance) velocity and $\tau$
as the total stress in the polymeric fluid. The FENE-CR viscoelastic fluid has
the advantage of being similarly easy to analyse using our spectral method, but
also including a finite extensibility for the polymers. This leads to a stress
which depends on a conformation tensor ($\mathbf{C}$) for the polymer dumbbells
via a non-linear spring force:
\begin{equation} 
    \mathbf{\tau} = \frac{1-\frac{3}{L^{2}}}{1 +
	\frac{L^{2}}{tr(\mathbf{C^{2}})}} \mathbf{C} \label{eq:KH_FENE_stress}
\end {equation}
Again, we used Gauss-Labatto points in the wall normal direction and Fourier
modes in the streamwise direction.

@@

At low Reynolds number, low $\beta$ and sufficiently large Weissenberg number
we observe an instability across a range of streamwise disturbance wavenumbers.
This instability persists for larger $\Delta$, albeit with a fairly complex
dependence on the boundaries. Before we examine the effect of the walls we will
consider the instability in the free shear limit.

\subsection{Free shear layer}

For a free shear layer we observe an instability grow from around $k \sim 0.06$
and move to lower wavenumber as the Weissenberg number increases. The
dispersion relations shown in figure \ref{fig:KH_inf_dispersions_vary_Wi} show
this dependence of the unstable eigenvalues on $\Wi$. The dispersion relation
for the instability remains broad in $k$ until $Wi \sim 100$ where a new
eigenvector becomes dominant.

@@

The height of the instability also increases with increasing Weissenberg
number. Extracting the peak growth rate from the dispersion relations at a
fixed $\Rey$ for various $\Wi$ allows us to show how the instability depends on
both $\beta$ and $\Wi$ (figure \ref{fig:KH_inf_disp_height_vary_Wi}). The
maximum growth rate saturates at $Wi \sim 50$ and remains approximately
constant with increasing Weissenberg number. The slight nonmonotonicity in the
curve is the result of the switchover in the dominant instability eigenmode.

@@

As the Reynold's number is reduced the Newtonian Kelvin-Helmholtz instability
disappears and an entirely separate, purely elastic instability arises.
Although the dominant eigenvalue changes often as $\Rey$ decreases,  the flow
is still unstable even down to $\Rey = 0.01$ (see figures
\ref{fig:KH_inf_disp_height_vary_Re} and \ref{fig:KH_dispersions_vary_Re}).
Again, the complicated dependence on $\Rey$ appears to be due to the appearance
and disappearance of new dominant eigenmodes of the instability.

@@

Now we use equations \ref{eq:KH_laminar_profile} for a two dimensional flow
with boundaries.  To validate our approach, we examine the high Reynolds number
regime in order to recover a Newtonian Kelvin-Helmholtz instability. At high
Reynolds number the instability is insensitive to the distance to the walls
when $\Delta =0.1$ (figure \ref{fig:KH_high_Re_walls_dependence}) and is
consistent with the Kelvin-Helmholtz instability given in \citet{Criminale2003}.

We see the maximum growth rate of the instability decrease with increasing
Weissenburg number, and this trend is stronger for smaller $\beta$ (figure
\ref{fig:KH_drag_reduction}). This is the same behaviour reported in
\cite{Azaiez1994} for the Kelvin-Helmholtz instability of a polymeric fluid.

@@

\begin{figure}[h]
%   \showthe\columnwidth %use this to determine size of figures.
    \centering \includegraphics[width=0.7\textwidth]{{KH_high_Re_vary_Wi}}
    \caption{ 
	Peak growth rate against Weissenberg number at $\Rey = 1000$, $\Delta =
	0.1$ The height of the dispersion relation is reduced for the
	Kelvin-Helmholtz instability as the Weissenberg number is increased.
	The grey dashed line is the height of the dispersion relation for the
	Newtonian fluid. The decrease in $\lambda_{r}$ corresponds to increased
	stability of the base flow, this matches the observations made in
	\cite{Azaiez1994}
    } 
    \label{fig:KH_drag_reduction} 
\end{figure}

At low Reynolds number, we can examine the purely elastic shear layer
instability of above. In this case find that the width of the channel has a
greater affect on the instablity of the flow. However, the flow is still
unstable at very low $\Delta$ (figure \ref{fig:KH_low_Re_walls_dependence}). At
$\Delta = 0.1$ the flow is unstable for $\Rey < 0.01$ far lower than the usual
threshold of $\Rey \sim 1000$ for a Newtonian Kelvin-Helmholtz instability
(figure \ref{fig:KH_walls_vary_Re}).

@@

As $\Delta$ decreases for a particular purely elastic instability, the width of
the instability increases slightly, however in the limit of very small $\Delta$
the eigenvector appears independent of the walls (figures
\ref{fig:KH_psi_low_delta} and \ref{fig:KH_psi_high_delta}). The extended
vortices in our purely elastic case seem to correspond to the eigenvectors in
\citet{Azaiez1994} at larger elasticity ($G \sim \Wi/\Rey$) . Their $G=5$ case would
correspond to only $\Wi=0.5$ for $\Rey=0.1$ in our units, such that the
instability they observe is still the inertial instability ( see
\citet{Azaiez1994} figure 8). Nonetheless at least one of their observations
does apply. The fact that the velocity disturbance is essentially zero in the
center of the channel, implies that the shear layer will be less likely to roll
up. In the Newtonian instability, net velocities within the shear layer will
distort its shape, however the purely elastic instability appears to have less
of an effect on the shear layer.

@@

As the Reynold's number is reduced the walls become more important to the
instability, this makes a true $\Rey = 0$ instability difficult for us to
observe. In order to remove the dependence on the walls and recover a free
shear instability it is necessary to move them further away. This problem gets
worse at lower $\Rey$.  By introducing just a very small contribution from
inertia, we can remove the dependence on the walls. This is partly why all of
our results are taken for $\Rey \lesssim 0.1$ on the length scale of the
instability.
